\mytitle

### 2012-02-21 papers/SAS2012/implicitization.tex
Rabottage, et correction de mise en forme suite à passage LLNCS

\usepackage{amsmath}
\usepackage{mathcomp,amsfonts,amssymb,mathabx}
\usepackage[mathscr]{euscript}
\usepackage{algorithmic,algorithm}
%\usepackage{graphicx}
%\usepackage{color}

@@

\usepackage{tikz}
\usetikzlibrary{arrows,automata}

@@

@@

\section{Bases}
\subsection{Static analysis by abstract interpretation}
\label{sec:static_analysis}
Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$. The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is the \emph{concrete domain}. An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion. In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
%
\footnote{%
Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents. This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}

@@

Widening operations have a somewhat counterintuitive behavior: the result of an analysis (postcondition) is not necessarily monotonic with respect to the precondition. For instance, classical interval analysis with widening and decreasing iterations on the following loop yields a loop invariant $x \in [0,+\infty)$ if the precondition is $x \in [0,0]$ , but $x \in [0,10]$ (strictly stronger) if the precondition is $x \in [0,10]$ (strictly weaker):
\lstinline|while (x != 10) x=x+1;|.
This means that, when using widenings, being locally more precise (for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP}) does not necessarily translate into higher final precision.
It is almost always possible to concoct examples where a supposedly more precise abstraction (e.g. polyhedra) yields less precise results than a less precise one (e.g. intervals), but are these examples representative of real programs?
This justifies our recourse to evaluation on realistic examples~(\S~\ref{sec:compare_techniques}).

@@

Most SMT solvers, including Z3,
%\footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
Yices,%
%\footnote{\url{http://yices.csl.sri.com/}}
and all those supporting the full SMTLIB2 standard \cite{BarST-SMTLIB},
offer an \emph{incremental} interface: the client program specifies the formula as an initially empty conjunction, to which additional constraints are added, and calls a ``check'' function answering whether it is satisfiable; 
it may then backtrack some of the constraints and add other ones without restarting from scratch.

\subsection{A simple, motivating example}
\label{subsec:rate_lim}
Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}, where \lstinline|input($a$, $b$)| stands for a nondeterministic input in $[a,b]$:
\lstinputlisting{rlim_int.c}
This program implements a construct commonly found in control programs (in e.g.
automotive or avionics): a rate limiter. For the sake of simplicity, we chose it
to be fed on a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and its input connected to a complex system with unknown output range.

@@

\begin{figure}
\centering
\begin{minipage}[c]{.19\textwidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
                    semithick,font=\footnotesize]

	\node[PRstate] (n0) {$p$};
	\node[state] (n1) [below left of=n0] {};
	\node[state] (n2) [below right of=n0] {};
	\node[state] (n3) [below right of=n1] {};
	\node[state] (n4) [below left of=n3] {};
	\node[state] (n5) [below right of=n3] {};
	\node[state] (n6) [below right of=n4] {};
	\node[state] (n7) [below left of=n6] {};
	\node[state] (n8) [below right of=n6] {};
	\node[PRstate] (n9) [below right of=n7] {$q$};

  \path [transition] 
		(n0) edge              node {} (n1);
  \path [transition] 
		(n0) edge              node {} (n2);
  \path [transition] 
		(n1) edge              node {} (n3);
  \path [transition] 
		(n2) edge              node {} (n3);
  \path [transition] 
		(n3) edge              node {} (n4);
  \path [transition] 
		(n3) edge              node {} (n5);
  \path [transition] 
		(n4) edge              node {} (n6);
  \path [transition] 
		(n5) edge              node {} (n6);
  \path [transition] 
		(n6) edge              node {} (n7);
  \path [transition] 
		(n6) edge              node {} (n8);
  \path [transition] 
		(n7) edge              node {} (n9);
  \path [transition] 
		(n8) edge              node {} (n9);
\end{tikzpicture}
\end{minipage} 
$\Longrightarrow$
\begin{minipage}[c]{.19\textwidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
                    semithick,font=\footnotesize]

	\node[PRstate] (n0) {$p$};
	\node (n1) [below left of=n0] {};
	\node (n2) [below right of=n0] {};
	\node (n3) [below right of=n1] {};
	\node (n4) [below left of=n3] {};
	\node (n5) [below right of=n3] {};
	\node (n6) [below right of=n4] {};
	\node (n7) [below left of=n6] {};
	\node (n8) [below right of=n6] {};
	\node[PRstate] (n9) [below right of=n7] {$q$};

@@

\subsection{Guided static analysis}
\label{sec:guided}
\emph{Guided static analysis} was proposed by \cite{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
Consider the following program, taken from \cite{DBLP:conf/sas/GopanR07}:
\lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}

\begin{figure}
\begin{center}
\includegraphics[scale=0.75]{gopan_reps}
\end{center}
\caption{The invariant for Prog.~\ref{lst:gopan_reps}: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
\label{fig:gopan_reps_invariant} 
\end{figure}

Classical iterations on the domain of convex polyhedra \cite{CousotHalbwachs78,BagnaraHRZ05SCP} or octagons \cite{DBLP:journals/lisp/Mine06} start with $x = 0 \land x = 0$, then continue with $x = y \land 0 \leq x \leq 1$.
The widening operator extrapolates from these two iterations and yields $x = y \land x \geq 0$.
From there, the ``else'' branch at line 5 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the postcondition computed at line 9 is $x \geq 0 \land y = 0$.
Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron (Fig.~\ref{fig:gopan_reps_invariant}), is
\begin{equation}
y \leq x \land y \leq 102-x \land y \geq 0.\label{eqn:triangle}
\end{equation}
Intuitively, this disappointing result is obtained because widening extrapolates from the first iterations of the loop, but the loop has two different phases ($x \leq 50$ and $x > 50$) with different behaviors, thus the extrapolation from the first phase is not valid for the second.

@@

The set of paths is kept implicit by an SMT formula $\rho$ expressing
the semantics of the program, assuming that the transition semantics can be
expressed within a decidable theory. For an easy construction of $\rho$, 
we also assume that the program is expressed in SSA form, meaning that each
variable is only assigned once in the transition graph. This is not a
restriction, since there exist standard algorithms that transform a program into
an SSA format.

@@

In order to find focus paths, we solve an SMT formula which is satisfiable when
there exist a path starting at a point $p_i \in P_R$ in a state included in the
current invariant candidate $X_i$, and arriving at a point $p_j \in P_R$ in a
state outside $X_j$. In this case, we construct this path using the model and
update $X_j$. When $p_i = p_j$, meaning that the path is actually a self-loop,
we can apply a widening/narrowing sequence, or even compute the transitive
closure of the loop (or an approximation thereof, or its application to $X_i$)
using abstract acceleration~\cite{DBLP:conf/sas/GonnordH06}.

\section{Guided analysis over the paths}
\label{sec:guided_multigraph}

@@

\subsection{Ascending iterations by Path-focusing}
\label{subsec:ascending}

@@

If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
successor of $p_i$ in the expanded multigraph, and $X_i^s$ the abstract value
associated to $p_i$ :
$$f(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
 \neg X_j^s)$$

The difference with~\cite{Monniaux_Gonnord_SAS11} is that we do not
work on the entire transition graph but on a subset of it. Therefore we
conjoin the formula $f(p_i)$ with the actual set of working paths,
noted $P$, expressed as a Boolean formula, where the Boolean variables are the
\emph{reachability predicates} of the control points. We can easily construct
this formula from the binary decision diagram using dynamic programming, and
avoiding an exponentially sized formula. In other words, we force the SMT solver
to give us a path included in $P$.

\subsection{Adding new paths}
\label{subsec:addingpaths}

@@

% À changer si on décide de ne plus distinguer $X^s$ et $X^d$
Also, to prevent the algorithm from adding too many paths at a time,
and in particular to prevent the algorithm from trying to re-add the
same path infinitely many times, we use another abstract value
associated to the control point $p_j$, noted $X_j^d$, which is distinct
from $X_j^s$, and initialized to $X_j^s$ right before computing new paths. 
In the SMT formula, we associate to $p_j$ the abstract value $X_j^s$ when $p_j$
is the starting point of the path, and $X_j^d$ when it is its destination point.
We thus check the satisfiability of the formula
$f'(p_i)$, where:
$$f'(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
\neg X_j^d)$$

$X_j^d$ is updated when the point $p_j$ is the target of a new path.
This way, further SMT queries do not
compute other paths with the same source and destination if it is not
needed (because these new paths would not make $X_j^d$ grow, hence
would not be returned by the SMT solver).

@@

\subsection{Termination}
Termination of this algorithm is guaranteed, because:
\begin{itemize}
\item 
the subset of paths $P$ strictly increases at each loop iteration, and bounded by the finite set of
paths in the entire graph. 
\item the set $P'$ always verifies $P \cap P' = \emptyset$ by construction, 
which guarantees that $P'$
will eventually be empty after a finite number of loop iterations.
\end{itemize}

@@

We revise the rate limiter described in \ref{subsec:rate_lim}. In this example,
\emph{Path-focusing} works well because all the paths starting at the
loop header are actually self loops. In such a case, the technique performs a
widening/narrowing sequence or accelerates the loop, thus leading to a precise
invariant. However, in some cases, there also exist paths that are not
self loops, in which case \emph{Path-focusing} applies widening. 
This widening may induce unrecoverable loss of precision.

@@

We choose $P_R$ as the set of loop headers of the function, plus the initial
state. In this case, we have three elements in $P_R$.
We also unroll the loop once, in order to distinguish the paths that do not go
through the loop from the others:
\begin{lstlisting}[numbers=none]
if (wait()) {
   while (wait()) {}
}
\end{lstlisting}

@@

\begin{enumerate}
	\item The first subset of paths is depicted on Figure
		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
	\item
		We then compute the set $P'$ of paths that have to be added into the
		next subgraph. 
		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_5$ to
		itself, and that goes through the \emph{else} branch of each 
		\emph{if-then-else}, is $-10 \leq \lstinline|x_old| \leq 10$. This path
		is then added to our subgraph. Moreover, there is no other path whose
		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
		Additionally, the path that goes from $p_5$ to
		$p_{10}$, that goes through the \emph{else} branch of each 
		\emph{if-then-else} at lines 7,8 and 9, is also 
		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
		new subgraph. Again, there is no new path that goes to $p_{10}$ with an
		image outside $-10 \leq \lstinline|x_old| \leq 10$.
	\item
		We apply Path Focusing on this new subgraph (\ref{fig:example-graph}
		Step 2). When we have reached an inductive invariant, we narrow, and the
		result for both $p_5$ and $p_{10}$ is the polyhedron 
		$-10000 \leq \lstinline|x_old| \leq 10000$.
	\item Finally, we compute the new subgraph. The SMT-solver does not find any
		new path that makes the abstract values grow, and the algorithm
		terminates.
\end{enumerate}

@@

  \path [transition] 
		(n00) edge  node {$\lstinline|x_old| \gets 0$} (n0);
\end{tikzpicture}
\end{minipage} 
\begin{minipage}[c]{.28\textwidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
                    semithick,font=\footnotesize]

@@

  \path [transition] 
		(n00) edge  node {$\lstinline|x_old| \gets 0$} (n0);
  \path [transition] 
		(n0) edge  [loop left] node [left, xshift=0.2cm] {
		$\begin{array}{r}
			-10000 \leq \lstinline|x| \leq 10000 \\
			\lstinline|x_old|-10 \leq \lstinline|x| \leq \lstinline|x_old|+10 /\\
			\lstinline|x_old| \gets \lstinline|x|
		\end{array}$
		} (n0);
  \path [transition] 
		(n0) edge node [left, yshift=-0.2cm, xshift=0.2cm] {
		$\begin{array}{r}
			-10000 \leq \lstinline|x| \leq 10000 \\
			\lstinline|x_old|-10 \leq \lstinline|x| \leq \lstinline|x_old|+10 /\\
			\lstinline|x_old| \gets \lstinline|x|
		\end{array}$
		} (n1);
\end{tikzpicture}
\end{minipage}
\caption{Ascending sequence of subgraphs}
\end{figure}

@@

\section{Disjunctive invariants}
\label{sec:disjunctive}

@@

Our improvement consists in a modification of the SMT formula we solve in
\ref{sec:guided_multigraph}.
We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
that we can easily find in the model which abstract value of the
disjunction of the source point has
to be chosen to make the invariant of the destination grow.
The resulting formula that is given to the SMT solver is defined
by $g(p_i)$.
When the formula is satisfiable, we know that the index $j$ of the starting
disjunct that has to be chosen is the one for which the associate Boolean value
$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
$\sigma_i(j,k)$, thus know the index of the disjunct to join with.

\begin{equation*}
g(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg b_j^s 
 \wedge 
\displaystyle\bigvee_{1 \leq k \leq m_i} (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
\neg d_l)
\wedge
\displaystyle\bigvee_{j \in Succ(i)} 
(b_j^d \wedge \bigwedge_{1 \leq k \leq m_i} (\neg X_{j,k}))
\end{equation*}

In our algorithm, the initialization of the abstract values slightly differs from
algorithm \ref{algo:combined} line~\ref{alg=X-init}, since we now have to
initialize each disjunct. Line~\ref{alg=X-init} is then replaced by:
\begin{algorithm}[!h]
\begin{algorithmic}[1] 
\FORALL {$k \in \{1,..,m_i\} \setminus \{\delta_i\}$}
	\STATE $X_{i,k} \gets \perp$
\ENDFOR
\STATE $X_{i,\delta_i} \gets I_{p_i}$
\end{algorithmic}
\end{algorithm}

@@

The \emph{Update} function can classically assign to $X_{i,\sigma_i(j,k)}$ the
value $X_{i,\sigma_i(j,k)} \widening (X_{i,\sigma_i(j,k)} \sqcup
\tau_{i,k}(X_{i,j}))$, or can integrate the special treatment for self loops
proposed by \cite{Monniaux_Gonnord_SAS11}, with widening/narrowing sequence or
acceleration.

@@

\section{Experimental comparisons}
\label{sec:experiments}

@@

\subsection{Analysis algorithm}
\label{sec:analysis-algorithm}
For each program, we distinguish a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$. Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g. \cite{BourdonclePhd}.

>>>>

\usepackage{amsmath}
\usepackage{mathcomp,amsfonts,amssymb,mathabx}
\usepackage[mathscr]{euscript}
\usepackage{algorithmic,algorithm}
\usepackage{paralist}
\usepackage{graphicx}
%\usepackage{color}

@@

\usepackage{tikz}
\usetikzlibrary{arrows,automata,positioning}

@@

\newcommand{\avirer}[1]{ {\color{red} #1} }
\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
%\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
%\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
%\newcommand{\MM}[1]{}
%\newcommand{\JH}[1]{}
%\newcommand{\DM}[1]{}

@@

\section{Bases}
\subsection{Static Analysis by Abstract Interpretation}
\label{sec:static_analysis}
\MM{On peut peut-être réduire un peu vu que tout le monde connait à
  SAS ?}

Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$. The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is the \emph{concrete domain}. An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion. In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
%
\footnote{%
Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents. This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}

@@

Widening operations have a somewhat counterintuitive behavior: the
result of an analysis (postcondition) is not necessarily monotonic
with respect to the precondition. For instance, classical interval
analysis with widening and decreasing iterations on the loop
\lstinline|while (x != 10) x=x+1;| yields a loop invariant $x \in
[0,+\infty)$ if the precondition is $x \in [0,0]$, but $x \in [0,10]$
(strictly stronger) if the precondition is $x \in [0,10]$ (strictly
weaker).
This means that, when using widenings, being locally more precise (for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP}) does not necessarily translate into higher final precision.
It is almost always possible to concoct examples where a supposedly more precise abstraction (e.g. polyhedra) yields less precise results than a less precise one (e.g. intervals), but are these examples representative of real programs?
This justifies our recourse to evaluation on real, large programs~(\S~\ref{sec:compare_techniques}).

@@

Most SMT solvers, including Z3,
%\footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
Yices,
%\footnote{\url{http://yices.csl.sri.com/}}
and all those supporting the full SMTLIB2 standard \cite{BarST-SMTLIB},
offer an \emph{incremental} interface: the client program specifies the formula as an initially empty conjunction, to which additional constraints are added, and calls a ``check'' function answering whether it is satisfiable; 
it may then backtrack some of the constraints and add other ones without restarting from scratch.

\subsection{A Simple, Motivating Example}
\label{subsec:rate_lim}
Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}, where \lstinline|input($a$, $b$)| stands for a nondeterministic input in $[a,b]$:
\lstinputlisting{rlim_int.c}
This program implements a construct commonly found in control programs (in e.g.
automotive or avionics): a rate limiter.
\avirer{For the sake of simplicity, we chose it
to be fed on a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and its input connected to a complex system with unknown output range.}

@@

\begin{figure}
\centering
\begin{minipage}[c]{.21\textwidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=.2cm and .5cm,
                    semithick,font=\footnotesize]

	\node[PRstate] (n0) {$p$};
	\node[state] (n1) [below left=of n0] {};
	\node[state] (n2) [below right=of n0] {};
	\node[state] (n3) [below right=of n1] {};
	\node[state] (n4) [below left=of n3] {};
	\node[state] (n5) [below right=of n3] {};
	\node[state] (n6) [below right=of n4] {};
	\node[state] (n7) [below left=of n6] {};
	\node[state] (n8) [below right=of n6] {};
	\node[PRstate] (n9) [below right=of n7] {$q$};

  \path [transition] (n0) -- (n1);
  \path [transition] (n0) -- (n2);
  \path [transition] (n1) -- (n3);
  \path [transition] (n2) -- (n3);
  \path [transition] (n3) -- (n4);
  \path [transition] (n3) -- (n5);
  \path [transition] (n4) -- (n6);
  \path [transition] (n5) -- (n6);
  \path [transition] (n6) -- (n7);
  \path [transition] (n6) -- (n8);
  \path [transition] (n7) -- (n9);
  \path [transition] (n8) -- (n9);
\end{tikzpicture}
\end{minipage} 
$\Longrightarrow$
\begin{minipage}[c]{.19\textwidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=.2cm and .5cm,
                    semithick,font=\footnotesize]

	\node[PRstate] (n0) {$p$};
	\node (n1) [below left=of n0] {};
	\node (n2) [below right=of n0] {};
	\node (n3) [below right=of n1] {};
	\node (n4) [below left=of n3] {};
	\node (n5) [below right=of n3] {};
	\node (n6) [below right=of n4] {};
	\node (n7) [below left=of n6] {};
	\node (n8) [below right=of n6] {};
	\node[PRstate] (n9) [below right=of n7] {$q$};

@@

\subsection{Guided Static Analysis}
\label{sec:guided}
\emph{Guided static analysis} was proposed by \cite{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
Consider the program in Fig.~\ref{fig:gopan_reps_invariant}, taken from \cite{DBLP:conf/sas/GopanR07}.

\begin{figure}
  \begin{minipage}{.3\linewidth}
    \lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}
  \end{minipage}
\hfill
\begin{minipage}{.5\linewidth}
\includegraphics[scale=0.75]{gopan_reps}
\vspace{-1ex}
\begin{equation}
y \leq x \;\land\; y \leq 102-x \;\land\; y \geq 0.\label{eqn:triangle}
\end{equation}
\end{minipage}
\caption{Example program and its invariant: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
\label{fig:gopan_reps_invariant} 
\end{figure}

Classical iterations on the domain of convex polyhedra \cite{CousotHalbwachs78,BagnaraHRZ05SCP} or octagons \cite{DBLP:journals/lisp/Mine06} start with $x = 0 \land x = 0$, then continue with $x = y \land 0 \leq x \leq 1$.
The widening operator extrapolates from these two iterations and yields $x = y \land x \geq 0$.
From there, the ``else'' branch at line 5 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the postcondition computed at line 9 is $x \geq 0 \land y = 0$.
Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron (Fig.~\ref{fig:gopan_reps_invariant}).

Intuitively, this disappointing result is obtained because widening extrapolates from the first iterations of the loop, but the loop has two different phases ($x \leq 50$ and $x > 50$) with different behaviors, thus the extrapolation from the first phase is not valid for the second.

@@

The set of paths is kept implicit by an SMT formula $\rho$ expressing
the semantics of the program, assuming that the transition semantics can be
expressed within a decidable theory. For an easy construction of $\rho$, 
we also assume that the program is expressed in SSA form, meaning that each
variable is only assigned once in the transition graph. This is not a
restriction, since there exists standard algorithms that transform a program into
an SSA format.

@@

In order to find focus paths, we solve an SMT formula which is satisfiable when
there exists a path starting at a point $p_i \in P_R$ in a state included in the
current invariant candidate $X_i$, and arriving at a point $p_j \in P_R$ in a
state outside $X_j$. In this case, we construct this path using the model and
update $X_j$. When $p_i = p_j$, meaning that the path is actually a self-loop,
we can apply a widening/narrowing sequence, or even compute the transitive
closure of the loop (or an approximation thereof, or its application to $X_i$)
using abstract acceleration~\cite{DBLP:conf/sas/GonnordH06}.

\section{Guided Analysis over the Paths}
\label{sec:guided_multigraph}

@@

\subsection{Ascending Iterations by Path-focusing}
\label{subsec:ascending}

@@

If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
successor of $p_i$ in the expanded multigraph, and $X_i^s$ the abstract value
associated to $p_i$ :
\[
f(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
 \neg X_j^s)
\]
The difference with~\cite{Monniaux_Gonnord_SAS11} is that we do not
work on the entire transition graph but on a subset of it. Therefore we
conjoin the formula $f(p_i)$ with the actual set of working paths,
noted $P$, expressed as a Boolean formula, where the Boolean variables are the
\emph{reachability predicates} of the control points. We can easily construct
this formula from the binary decision diagram using dynamic programming, and
avoiding an exponentially sized formula. In other words, we force the SMT solver
to give us a path included in $P$.

\subsection{Adding New Paths}
\label{subsec:addingpaths}

@@

% À changer si on décide de ne plus distinguer $X^s$ et $X^d$
Also, to prevent the algorithm from adding too many paths at a time,
and in particular to prevent the algorithm from trying to re-add the
same path infinitely many times, we use another abstract value
associated to the control point $p_j$, noted $X_j^d$, which is distinct
from $X_j^s$, and initialized to $X_j^s$ right before computing new paths. 
In the SMT formula, we associate to $p_j$ the abstract value $X_j^s$ when $p_j$
is the starting point of the path, and $X_j^d$ when it is its destination point.
We thus check the satisfiability of the formula
$f'(p_i)$, where:
\[
f'(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
\neg X_j^d)
\]
$X_j^d$ is updated when the point $p_j$ is the target of a new path.
This way, further SMT queries do not
compute other paths with the same source and destination if it is not
needed (because these new paths would not make $X_j^d$ grow, hence
would not be returned by the SMT solver).

@@

\subsection{Termination}
Termination of this algorithm is guaranteed, because:
\begin{inparaenum}
\item 
the subset of paths $P$ strictly increases at each loop iteration, and bounded by the finite set of
paths in the entire graph. 
\item the set $P'$ always verifies $P \cap P' = \emptyset$ by construction, 
which guarantees that $P'$
will eventually be empty after a finite number of loop iterations.
\end{inparaenum}

@@

We revise the rate limiter described in \ref{subsec:rate_lim}. In this example,
\emph{Path-focusing} works well because all the paths starting at the
loop header are actually self loops. In such a case, the technique performs a
widening/narrowing sequence or accelerates the loop, thus leading to a precise
invariant. However, in some cases, there also exists paths that are not
self loops, in which case \emph{Path-focusing} applies widening. 
This widening may induce unrecoverable loss of precision.

@@

We choose $P_R$ as the set of loop headers of the function, plus the initial
state. In this case, we have three elements in $P_R$.
We also unroll the loop once, in order to distinguish the paths that do not go
through the loop from the others:
\lstinline|if (wait()) {while (wait()) {}}|

@@

\begin{enumerate}
	\item The first subset of paths is depicted on Figure
		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
	\item
		We then compute the set $P'$ of paths that have to be added into the
		next subgraph. 
		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_5$ to
		itself, and that goes through the \emph{else} branch of each 
		\emph{if-then-else}, is $-10 \leq \lstinline|x_old| \leq 10$. This path
		is then added to our subgraph. Moreover, there is no other path whose
		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
		Additionally, the path that goes from $p_5$ to
		$p_{10}$, that goes through the \emph{else} branch of each 
		\emph{if-then-else} at lines 7, 8 and 9, is also 
		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
		new subgraph. Again, there is no new path that goes to $p_{10}$ with an
		image outside $-10 \leq \lstinline|x_old| \leq 10$.
	\item
		We apply Path Focusing on this new subgraph (\ref{fig:example-graph}
		Step 2). When we have reached an inductive invariant, we narrow, and the
		result for both $p_5$ and $p_{10}$ is the polyhedron 
		$-10000 \leq \lstinline|x_old| \leq 10000$.
	\item Finally, we compute the new subgraph. The SMT-solver does not find any
		new path that makes the abstract values grow, and the algorithm
		terminates.
\end{enumerate}

@@

  \path [transition] 
		(n00) edge  node {$\lstinline|x_old| \gets 0$} (n0);
\end{tikzpicture}
\end{minipage} 
\begin{minipage}[c]{.75\textwidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
                    semithick,font=\footnotesize]

@@

  \path [transition] 
		(n00) edge  node {$\lstinline|x_old| \gets 0$} (n0);
  \path [transition] 
		(n0) edge  [loop left] node [left, xshift=0.2cm] {
		$\begin{array}{r}
			-10000 \leq \lstinline|x| \leq 10000 \\
			\lstinline|x_old|-10 \leq \lstinline|x| \leq \lstinline|x_old|+10 /\\
			\lstinline|x_old| \gets \lstinline|x|
		\end{array}$
		} (n0);
  \path [transition] 
		(n0) edge node [left, yshift=-0.2cm, xshift=0] {
		$\begin{array}{r}
			-10000 \leq \lstinline|x| \leq 10000 \\
			\lstinline|x_old|-10 \leq \lstinline|x| \leq \lstinline|x_old|+10 /\\
			\lstinline|x_old| \gets \lstinline|x|
		\end{array}$
		} (n1);
\end{tikzpicture}
\end{minipage}
\caption{Ascending sequence of subgraphs}
\end{figure}

@@

\section{Disjunctive Invariants}
\label{sec:disjunctive}

@@

Our improvement consists in a modification of the SMT formula we solve in
\ref{sec:guided_multigraph}.
We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
that we can easily find in the model which abstract value of the
disjunction of the source point has
to be chosen to make the invariant of the destination grow.
The resulting formula that is given to the SMT solver is defined
by $g(p_i)$.
When the formula is satisfiable, we know that the index $j$ of the starting
disjunct that has to be chosen is the one for which the associate Boolean value
$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
$\sigma_i(j,k)$, thus know the index of the disjunct to join with.
%
\begin{equation*}
g(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg b_j^s 
 \wedge 
\displaystyle\bigvee_{1 \leq k \leq m_i}\!\! (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
\neg d_l)
\wedge
\displaystyle\bigvee_{j \in Succ(i)}\!\!
(b_j^d \wedge \bigwedge_{1 \leq k \leq m_i}\!\! (\neg X_{j,k}))
\end{equation*}
%
In our algorithm, the initialization of the abstract values slightly differs from
algorithm \ref{algo:combined} line~\ref{alg=X-init}, since we now have to
initialize each disjunct. Instead of Line~\ref{alg=X-init}, we initialize $X_{i,k}$ with $\perp$ for all
$k \in \{1,..,m_i\} \setminus \{\delta_i\}$, and $X_{i,\delta_i}$ with
$\gets I_{p_i}$.

% MM: pas la place pour un vrai algo => inliné dans le texte.
% Line~\ref{alg=X-init} is then replaced by:
% %\begin{algorithm}[!h]
% \begin{algorithmic}[1] 
% \FORALL {$k \in \{1,..,m_i\} \setminus \{\delta_i\}$}
% 	\STATE $X_{i,k} \gets \perp$
% \ENDFOR
% \STATE $X_{i,\delta_i} \gets I_{p_i}$
% \end{algorithmic}
% %\end{algorithm}

@@

The \emph{Update} function can classically assign  $X_{i,\sigma_i(j,k)} \widening (X_{i,\sigma_i(j,k)} \sqcup
\tau_{i,k}(X_{i,j}))$ to $X_{i,\sigma_i(j,k)}$, or can integrate the special treatment for self loops
proposed by \cite{Monniaux_Gonnord_SAS11}, with widening/narrowing sequence or
acceleration.

@@

\section{Implementation and Experimental Comparisons}
\label{sec:experiments}

@@

\subsection{Analysis Algorithm}
\label{sec:analysis-algorithm}
For each program, we distinguish a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$. Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g. \cite{BourdonclePhd}.
